---
title: "Impact of Severe Weather Events on Health and Economy"
author: "John McKinzie"
date: "01/18/2015"
output: html_document
---

# Synopsis

This analysis uses [data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) from the 
[U.S. National Oceanic and Atmospheric Administration's (NOAA) storm events database](http://www.ncdc.noaa.gov/stormevents/).
The objective was to determine which storms events, from 1950 to November 2011, caused the most fatalities, injuries,
property damage and crop damage. Observations with missing or undetermined values were either removed of transformed.
Our findings show that the **TORNADO** event is by the far the most dangerous weather event for both fatatlies and 
injures. The **TORNADO** event's combined number were more than 10x the **EXCESSIVE HEAT** event, which 
came in second. For property damage, the **FLOOD** event was by far the highest, while **EXCESSIVE WETNESS** was 
responsible for the most crop damage.

# Data Processing

For the fatality and injury data, NA values were replace with 0, as not to affect the totals. It was done if this manner
as opposed removing the observations, so that if one of the variables had an NA, the other value could still be used for 
the total calculations. The fatality and injury data was then grouped by event type and summed. An additional calculated
column for the combination of fatalties and injuries was added.

Property and crop damage in the data set was stored in scientific notation. This meant there was a column for the 
coefficient and exponent for each damage type. Coeffection values that were NA were converted to zero. In addition,
expontent values were either converted to zero if NA or and unknown type. If the coeffection was a character 
respresentation of an exponent value (e.g. M = million = 6), it was converted to its corresponding numeric value. The 
property and crop damage data was then grouped by event type and summed. Two additional calculated columns were added 
that contained the decimal value represtation the value. This value as scaled ot represent millions. An additional 
calculated column for the combination of property and property damage was added.

```{r fatalities_injuries, message=FALSE}
temp <- tempfile()
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", temp)
storm_data <- read.csv(bzfile(temp))
unlink(temp)

library(dplyr)

# set number of top events to plot
top_limit <- 5

# fatalies or injuries data
storm_data_by_event <- storm_data

# convert NAs to zero
storm_data_by_event$FATALITIES[is.na(storm_data$FATALITIES)] <- 0
storm_data_by_event$INJURIES[is.na(storm_data$INJURIES)] <- 0

# group by event
storm_data_by_event <- storm_data_by_event %>%
  group_by(EVTYPE) %>%
    summarise(FATALITIES_TOTAL = sum(FATALITIES), INJURIES_TOTAL = sum(INJURIES), BOTH_TOTAL = sum(FATALITIES + INJURIES))

# property and crop damage data
storm_data_by_impact <- storm_data
# complete_rows <- complete.cases(storm_data_by_impact[, c("PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")])
# storm_data_by_impact <- storm_data_by_impact[complete_rows, ]
# storm_data_by_impact <- storm_data_by_impact[(storm_data_by_impact$PROPDMG > 0 & storm_data_by_impact$CROPDMG > 0), ]

# convert levels to characters
storm_data_by_impact$PROPDMGEXP <- as.character(storm_data_by_impact$PROPDMGEXP)
storm_data_by_impact$CROPDMGEXP <- as.character(storm_data_by_impact$CROPDMGEXP)

# convert NAs or non-numeric exponent values to numeric
storm_data_by_impact$PROPDMG[(is.na(storm_data_by_impact$PROPDMG) | 
                              storm_data_by_impact$PROPDMGEXP %in% c("", "-", "?", "+"))] <- 0  
storm_data_by_impact$PROPDMGEXP[(is.na(storm_data_by_impact$PROPDMGEXP) | 
                                 storm_data_by_impact$PROPDMGEXP %in% c("", "-", "?", "+"))] <- 0  
storm_data_by_impact$PROPDMGEXP[storm_data_by_impact$PROPDMGEXP %in% c("k", "K")] <- 3  
storm_data_by_impact$PROPDMGEXP[storm_data_by_impact$PROPDMGEXP %in% c("m", "M")] <- 6  
storm_data_by_impact$PROPDMGEXP[storm_data_by_impact$PROPDMGEXP %in% c("b", "B")] <- 9  

storm_data_by_impact$CROPDMG[(is.na(storm_data_by_impact$CROPDMG) | 
                              storm_data_by_impact$CROPDMGEXP %in% c("", "-", "?", "+"))] <- 0  
storm_data_by_impact$CROPDMGEXP[(is.na(storm_data_by_impact$CROPDMGEXP) | 
                                 storm_data_by_impact$PROPDMGEXP %in% c("", "-", "?", "+"))] <- 0  
storm_data_by_impact$CROPDMGEXP[storm_data_by_impact$CROPDMGEXP %in% c("k", "K")] <- 3  
storm_data_by_impact$CROPDMGEXP[storm_data_by_impact$CROPDMGEXP %in% c("m", "M")] <- 6  
storm_data_by_impact$CROPDMGEXP[storm_data_by_impact$CROPDMGEXP %in% c("b", "B")] <- 9  

# group by event type and conver to millions
storm_data_by_impact <- storm_data_by_impact %>%
  group_by(EVTYPE, PROPDMGEXP, CROPDMGEXP) %>%
  summarise(PROPDMG_UNSCALED_TOTAL = sum(PROPDMG) / 1000000, 
            CROPDMG_UNSCALED_TOTAL = sum(CROPDMG) / 1000000)

# scale base values by exponent
storm_data_by_impact$PROPDMG_SCALED_TOTAL <- suppressWarnings(storm_data_by_impact$PROPDMG_UNSCALED_TOTAL * (10 ^ as.integer(storm_data_by_impact$PROPDMGEXP)))
storm_data_by_impact$CROPDMG_SCALED_TOTAL <- suppressWarnings(storm_data_by_impact$CROPDMG_UNSCALED_TOTAL * (10 ^ as.integer(storm_data_by_impact$CROPDMGEXP)))

# group again by event type
storm_data_by_impact <- storm_data_by_impact %>%
  group_by(EVTYPE) %>%
  summarise(PROPDMG_TOTAL = sum(PROPDMG_SCALED_TOTAL), 
            CROPDMG_TOTAL = sum(CROPDMG_SCALED_TOTAL))

# convert NAs to 0 so as not to affect totals
storm_data_by_impact$PROPDMG_TOTAL[is.na(storm_data_by_impact$PROPDMG_TOTAL)] <- 0
storm_data_by_impact$CROPDMG_TOTAL[is.na(storm_data_by_impact$CROPDMG_TOTAL)] <- 0
storm_data_by_impact$BOTH_TOTAL = storm_data_by_impact$PROPDMG_TOTAL + storm_data_by_impact$CROPDMG_TOTAL
```

# Results

The approach of the analysis was to rank weather the top `r top_limit` weather events that caused the most fatalities, 
injuries, property damage and crop damage. By doing this, decision makers will have additional efforst to support
their planning efforts.

## Fatalities and Injuries

The following plots show the top `r top_limit` fatility, injury and combined events for the past 60+ year. For 
fatalities only, the **TORNADO** event is the most deadly with over 5500 fatalities. This is more than double **EXCESSIVE HEAT**, 
which is the second-most fatal event.

Again **TORNADO** events are the most dangerous with regards to injuries caused, with over 80,000. The **TSTM WIND**
(thunderstorms), **FLOOD** and**EXCESSIVE HEAT** events follow, all with less than 10,000 injuries.

When combine both fatalities and injuries are combined, the **TORNADO** events in not surprinsly first. The **TORNADO** event's 
combined numbers approach 100,000, more than 10x the **EXCESSIVE HEAT** event, which is a distant second.

```{r fatality_injury_plots, message=FALSE, fig.height=24, fig.width=8}
library(ggplot2)
library(grid)
library(gridExtra)

# plot fatalities
storm_data_by_event_fatalities <- arrange(storm_data_by_event, desc(FATALITIES_TOTAL))[1:top_limit, ]
p <- with(storm_data_by_event_fatalities, 
          qplot(reorder(EVTYPE, -FATALITIES_TOTAL), 
                FATALITIES_TOTAL,
                main = paste("Top", top_limit, "Fatality-Causing Weather Events"), 
                xlab = "Event Type", 
                ylab = "Total Fatalities") + 
          geom_bar(stat="identity") + 
          theme(axis.text.x = element_text(angle = 90)))

# rm(storm_data_by_event_fatalities)

# plot injuries
storm_data_by_event_injuries <- arrange(storm_data_by_event, desc(INJURIES_TOTAL))[1:top_limit, ]
q <- with(storm_data_by_event_injuries, 
          qplot(reorder(EVTYPE, -INJURIES_TOTAL), 
                INJURIES_TOTAL, 
                main = paste("Top", top_limit, "Injury-Causing Weather Events"),
                xlab = "Event Type", 
                ylab = "Total Injuries") + 
          geom_bar(stat="identity") + 
          theme(axis.text.x = element_text(angle = 90)))

# rm(storm_data_by_event_injuries)

# plot both
storm_data_by_event_total <- arrange(storm_data_by_event, desc(BOTH_TOTAL))[1:top_limit, ]
r <- with(storm_data_by_event_total,
          qplot(reorder(EVTYPE, -BOTH_TOTAL),
                BOTH_TOTAL,
                main = paste("Top", top_limit, "Fatality and Injury-Causing Weather Events"),
                xlab = "Event Type", 
                ylab = "Total Fatalities and Injuries") + 
          geom_bar(stat="identity") + 
          theme(axis.text.x = element_text(angle = 90)))

# rm(storm_data_by_event_total)

grid.arrange(p, q, r)
```

## Property and Crop Damage

For property damange, the **FLOOD** event was number one, with property damage approaching $125 billion. The 
**HURRICANE/TYPHOON** event was second, with significant damage totaling just over $25 billion. In third place was the
**HURRICANE**. This event could probably be combined with with the **HURRICANE/TYPHOON** event.

When looking at crop damage, the **EXCESSIVE WETNESS** (see NOAA documentation for event definitions) event was first 
with just over $100 million in damage. A close second was the **ICE STORM** with roughly $5 billion also. The **FLOOD**
event, which had the most property damage, had the third most crop damage.

Combining the damages caused to both property and crops, the top three, **FLOOD**, **HURRICANE/TYPHOON** and **TORNADO**,  
were the same as for just property. This is due to the fact that property damage values were a couple orders of magnitude 
larger than the crop damage.

```{r property_crop_plots, fig.height=24, fig.width=8}
# plot property
storm_data_by_impact_propdmg <- arrange(storm_data_by_impact, desc(PROPDMG_TOTAL))[1:top_limit, ]
t <- with(storm_data_by_impact_propdmg, 
          qplot(reorder(EVTYPE, -PROPDMG_TOTAL), 
                PROPDMG_TOTAL,
                main = paste("Top", top_limit, "Property Damage-Causing Weather Events"),
                xlab = "", 
                ylab = "Total Property (millions of $)") + 
          geom_bar(stat="identity") + 
          theme(axis.text.x = element_text(angle = 90)))

# rm(storm_data_by_impact_propdmg)

# plot crop
storm_data_by_impact_cropdmg <- arrange(storm_data_by_impact, desc(CROPDMG_TOTAL))[1:top_limit, ]
u <- with(storm_data_by_impact_cropdmg, 
          qplot(reorder(EVTYPE, -CROPDMG_TOTAL), 
                CROPDMG_TOTAL, 
                main = paste("Top", top_limit, "Crop Damage-Causing Weather Events"),
                xlab = "", 
                ylab = "Total Crop (millions of $)") + 
          geom_bar(stat="identity") + 
          theme(axis.text.x = element_text(angle = 90)))

# rm(storm_data_by_impact_cropdmg)

# plot both
storm_data_by_impact_both <- arrange(storm_data_by_impact, desc(BOTH_TOTAL))[1:top_limit, ]
v <- with(storm_data_by_impact_both,
          qplot(reorder(EVTYPE, -BOTH_TOTAL),
                BOTH_TOTAL,
                main = paste("Top", top_limit, "Property and Crop Damage-Causing Weather Events"),
                xlab = "", 
                ylab = "Total Property and Crops (millions of $)") + 
          geom_bar(stat="identity") + 
          theme(axis.text.x = element_text(angle = 90)))

# rm(storm_data_by_impact_both)

grid.arrange(t, u, v)
```
